Article Title

### 2014-07-14 plos.tex
minor edits

\begin{figure}    \floatbox[{\capbeside\thisfloatsetup{capbesideposition={right,center},capbesidewidth=8.3cm}}]{figure}[\FBwidth]
    {\caption{{\bf Acquisition of neural signals for the estimation of noise correlations.}
    Visual stimuli comprising full-field drifting gratings interleaved with blank screens ({\bf A}) presented during two-photon recordings of somatic calcium signals using fast 3D random-access microscopy ({\bf B}). 
    {\bf C--F.} Calcium activity data from an example site.
    {\bf C.} Representative calcium signals of seven cells, downsampled to 20 Hz, out of the 292 total recorded cells. Spiking activity inferred by non-negative deconvolution is shown by red ticks below the trace.
    {\bf D.} The spatial arrangement and orientation tuning of the 292 cells from the imaged site. The cells' colors indicate their orientation preferences. The gray cells were not significantly tuned.
    {\bf E.} The sample noise correlation matrix of the activity of the neural population. 
    {\bf F.} Histogram of noise correlation coefficients in one site. The red line indicates the mean correlation coefficient of 0.020.
} \label{fig:2}}
    {\includegraphics[width=8.3cm]{figures/Figure02.pdf}}
\end{figure}

@@

\paragraph{Surgery and two-photon imaging.}
The surgical procedures and data acquisition were performed as described in \cite{Cotton:2013}. Briefly, C57BL/6J mice (aged p40--60) were used. Anesthesia was initiated with isoflurane (3\%) and maintained with the mixture of fentanyl (0.05 mg/kg), midazolam (5 mg/kg), and medetomidine (0.5 mg/kg), with boosts of half the initial dose every 3 hours.  A craniotomy was performed over the right primary visual cortex.  Membrane-permeant calcium indicator Oregon Green 488 BAPTA-1 AM (OGB-1, Invitrogen) was loaded by bolus injection.  The craniotomy was sealed using a glass coverslip secured with dental cement.

@@

\paragraph{Visual stimulus.}
The visual stimulus consisted of full-field drifting gratings with 90\% contrast, luminance of 10 cd/m$^2$, spatial frequency of 0.08 cycles/degree, and temporal frequency of 2 cycles/s. Two sets of stimuli were presented for each imaging site: the first to map directional tuning and the second to estimate noise correlations. Directional tuning was mapped using a pseudo-random sequence of drifting gratings at sixteen directions of motion, 500 ms per direction, without blanks, with 12--30 trials for each direction of motion. The data for covariance estimation were collected during presentations of full-field drifting gratings with the same parameters as those used in directional tuning except only two directions (in 9 datasets) or five directions (in 22 datasets) were used and the presentations lasted 1 second and were separated by 1-second blanks.  Each stimulus condition was presented between 100 and 300 times.

@@

The measured fluorescent traces were deconvolved to reconstruct the firing rates for each neuron: First, the first principal component was subtracted from the traces, which reduced common mode noise related to small cardiovascular movements \cite{Cotton:2013}. The resulting traces were high-pass filtered above 0.1 Hz and downsampled to 20 Hz (\figref{2}{C}). Then, the firing rates were estimated using a fast non-negative deconvolution algorithm \cite{Vogelstein:2010}.

Orientation tuning was computed by fitting the mean firing rates in response to gratings drifting in directions $\phi$ with two-peaked von Mises tuning functions of the form $f(\phi)=a + b\exp\left[\frac 1 w(\cos(\phi-\theta)-1) \right] + c\exp\left[\frac 1 w(\cos(\phi-\theta+\pi)-1) \right]$ where $b\ge c$ are the amplitudes of the two respective peaks, $w$ is the tuning width, and  $\theta$ is the preferred direction. The significance of the fit was determined by the permutation test: the labels of the direction were randomly permuted 10,000 times; the $p$-values of the fits were computed as the fraction of the permuted datasets for which the $R^2$ value of the tuning function fit equaled or exceeded that of the real data.  Cells were considered tuned with $p$-values below 0.05.

@@

\paragraph{Cross-validation.}
To compare the performance of the estimators, we used conventional 10-fold cross-validation: Trials were randomly divided into 10 subsets with approximately equal numbers of trials of each condition in each subset. Each subset was then used as the testing sample with the rest of the data used as the training sample for estimating the covariance matrix. The resulting validation losses were averaged to produce the \emph{validation loss} reported throughout the paper.

@@

When the validation loss was not required, only the inner loop of cross-validation was used on the entire dataset.  This approach was used to compute the covariance matrix estimates and their excess loss in the simulation study (\figref{1}{\,Rows 4 and 5}) and to analyze the partial correlation structure of the $C_{\sf sparse+latent}$ estimator (Fig.~\ref{fig:4}--\ref{fig:6}) once its superior performance has been established.

@@

Estimator $C_{\sf diag}$ (Eq.~\ref{eq:c-diag})  used two hyperparameters: the covariance shrinkage intensity $\lambda \in [0,1]$ and variance shrinkage intensity $\alpha \in [0,1]$.  The variances (the diagonal of $C_{\sf sample}$) were shrunk toward (linearly mixed with) their mean value:
\begin{equation}
D = (1-\alpha)C_{\sf sample}\circ I + \alpha \frac 1 p \Tr(C_{\sf sample}) I
\end{equation}
Then the diagonal matrix $D$ was used as the target of covariance shrinkage (Eq.~\ref{eq:c-diag}) to produce the final regularized estimate.

In estimator $C_{\sf factor}$ (Eq.~\ref{eq:c-factor}), the low-rank matrix $L$ and the diagonal matrix $D$ are found by solving the minimization problem
\begin{equation}
(L,D) = \argmin\limits_{\hat L,\hat D} \loss{\hat L + \hat D,C_{\sf sample}},
\end{equation}
by an expectation-maximization (EM) algorithm.  The final regularized estimate is obtained by linear shrinkage of the diagonal matrix $D$ toward its mean diagonal value.

Estimator $C_{\sf sparse}$  has one hyperparameter $\lambda$ to regulate the sparsity of its precision matrix $S$. The precision matrix is obtained by minimizing the $L_1$-penalized loss:
\begin{equation}
S = \argmin\limits_{\hat S \succ 0} \loss{{\hat S}^{-1},C_{\sf sample}} + \lambda \|\hat S \|_1
\end{equation}
where $\hat S\succ 0$ denotes the constraint that $\hat S$ be a positive definite matrix and $\|\hat S\|_1$ is the element-wise $L_1$ norm of the matrix $\hat S$. This problem formulation is known as \emph{graphical lasso} \cite{Meinshausen:2006, Friedman:2008}. To solve this minimization problem, we adapted the alternative-direction method of multipliers (ADMM) \cite{Ma:2013}. 
Then the final estimate is the inverse of $S$ (Eq.~\ref{eq:c-sparse}). Unlike $C_{\sf diag}$ and $C_{\sf factor}$, this estimator does not include linear shrinkage: the selection of the sparsity level provides sufficient flexibility to fine tune the regularization level.

Estimator $C_{\sf sparse+latent}$ has two hyperparameters: the number of latent units $d$ and sparsity parameter $\lambda$. $C_{\sf sparse+latent}$ estimates a larger sparse precision matrix $S^\ast$ of the joint distribution of the $p$ observed neurons and $d$ latent units.  
\begin{equation}
S^\ast=
\begin{pmatrix}
S & S_{12} \\
S_{12}^\T & S_{22}
\end{pmatrix},
\end{equation}
where the $p\times p$ partition $S$ corresponds to the visible units and expresses their partial correlation structure, and $S_{12}$ and $S_{22}$ are of size $p\times d$ and $d\times d$, respectively. 
Then the covariance matrix of the observed population is 
\begin{equation}
C_{\sf sparse+latent} = \left(S-S_{12}S_{22}^{-1}S_{12}^\T\right)^{-1}
\end{equation}
The $p\times p$  matrix $L=S_{12}S_{22}^{-1}S_{12}^\T$ has rank $d$. Rather than searching for the optimal sparse structure of $S_{12}$ and $S_{22}$, an ill-posed problem, we estimated these components together as the low-rank matrix $L$. We solved this problem by adapting the ADMM algorithm to minimizing the loss function with penalty on the $L_1$-norm  of $S$ and the nuclear norm of $L$ \cite{Chandrasekaran:2010,Ma:2013}:
\begin{equation}\label{eq:ma}
(S,L) = \argmin\limits_{\hat S,\hat L} \loss{\hat S-\hat L, C_{\sf sample}} + \alpha\|\hat S\|_1 + \beta\Tr(\hat L)
\end{equation}

The partial correlation matrix with both sparse and low-rank components is computed from $C_{\sf sparse+latent}$ according to Eq.~\ref{eq:partial}; it includes the effects of interactions between the visible and latent units.  This estimate of the partial correlations was used in analyses of average partial correlations (\figref{4}{B and E} and \figref{5}{D}, and \figref{6}{A--C}).  The sparse component $S$ was used separately in analyses of the connectivity graphs (\figref{4}{F, G, H} and \figref{5}{B, C, E, and F} and \figref{6}{D--F}); in this case, the partial correlations were computed by normalizing the sparse component alone without the low-rank component: 
\begin{equation}
P_{\sf sparse} = -(S\circ I)^{-\frac 1 2} S  (S\circ I)^{-\frac 1 2}
\end{equation}

@@

\paragraph{Cross-validation with condition-specific variances.}
This study required an unbiased estimate of the discrepancy between various correlation matrix estimates and the true value of the correlation matrix. This was accomplished by cross-validation whereby regularized covariance matrix estimates computed from training samples were compared to unbiased covariance matrix estimates computed from independent testing samples. Loss functions were defined with respect to the covariance matrix rather than the correlation matrix because no unbiased estimate of the correlation matrix exists for finite sample sizes in the general case \cite{Fisher:1921}. In particular, we showed that with the normal loss function (Eq.~\ref{eq:loss}), the validation loss $\loss{C,C_{\sf sample}^\prime}$ is an unbiased estimate of true loss when $C_{\sf sample}^\prime$ is an unbiased estimate. However, this approach only worked under the assumption of uniform variances across all conditions included in the training and testing samples.  Here, we extend the validation loss to the estimation and evaluation of a common correlation matrix across multiple conditions with different variances:

Let $\{x(t):t\in T_c\}$ be a set of measurements collected in condition $c$ from the training sample and let $\{x(t):t \in T_c^\prime\}$ be a set of measurements in condition $c$ from the testing sample, where $T_c$ and $T_c^\prime$ are time bin indices for the training and testing samples, respectively. We marked all quantities estimated from the testing sample with a prime mark: \emph{e.g.}\,$C_{c,\sf sample}^\prime$, $n_c^\prime$.

@@

Since $\bar x_c= \frac 1 {n_c} \sum\limits_{t \in T_c}x(t)$ is estimated from the training sample in both estimates, $C_{c,\sf sample}^\prime$ is an unbiased estimate of the true covariance matrix, $\Sigma$, and can be used for validation whereas $C_{c,\sf sample}$ is biased toward zero and is used as input for regularized covariance estimates.

To apply regularization to the correlation matrix, we first reconstruct the common covariance matrix.
The common correlation matrix $R_{\sf sample}$ is estimated by averaging the condition-specific correlations:
\begin{equation}
	R_{\sf sample}  
	= \frac 1 n \sum\limits_c n_c \left(V_{c,\sf sample}^{-\frac 1 2}C_{c,\sf sample}V_{c,\sf sample}^{-\frac 1 2}\right)
        = \frac 1 n \sum\limits_c \sum\limits_{t \in T_c} z(t)z(t)^\T,
\end{equation}
where $n=\sum\limits_c n_c$ and $V_{c,\sf sample} = C_{c,\sf sample} \circ I$ is the diagonal matrix containing the sample variances. Then $R_{\sf sample}$ is simply the covariance matrix of the $z$-score signal 
$z(t) = V_{c,\sf sample}^{-\frac 1 2} \left(x(t) - \bar x_c\right)$ of the training sample.

>>>>

\begin{figure}    \floatbox[{\capbeside\thisfloatsetup{capbesideposition={right,center},capbesidewidth=8.3cm}}]{figure}[\FBwidth]
    {\caption{{\bf Acquisition of neural signals for the estimation of noise correlations.}
    Visual stimuli comprising full-field drifting gratings interleaved with blank screens ({\bf A}) presented during two-photon recordings of somatic calcium signals using fast 3D random-access microscopy ({\bf B}). 
    {\bf C--F.} Calcium activity data from an example site.
    {\bf C.} Representative calcium signals of seven cells, downsampled to 20 Hz, out of the 292 total recorded cells. Spiking activity inferred by nonnegative deconvolution is shown by red ticks below the trace.
    {\bf D.} The spatial arrangement and orientation tuning of the 292 cells from the imaged site. The cells' colors indicate their orientation preferences. The gray cells were not significantly tuned.
    {\bf E.} The sample noise correlation matrix of the activity of the neural population. 
    {\bf F.} Histogram of noise correlation coefficients in one site. The red line indicates the mean correlation coefficient of 0.020.
} \label{fig:2}}
    {\includegraphics[width=8.3cm]{figures/Figure02.pdf}}
\end{figure}

@@

\paragraph{Surgery and two-photon imaging.}
The surgical procedures and data acquisition were performed as described in \cite{Cotton:2013}: C57BL/6J mice (aged p40--60) were used. Anesthesia was initiated with isoflurane (3\%) and maintained with the mixture of fentanyl (0.05 mg/kg), midazolam (5 mg/kg), and medetomidine (0.5 mg/kg), with boosts of half the initial dose every 3 hours.  A craniotomy was performed over the right primary visual cortex.  Membrane-permeant calcium indicator Oregon Green 488 BAPTA-1 AM (OGB-1, Invitrogen) was loaded by bolus injection.  The craniotomy was sealed using a glass coverslip secured with dental cement.

@@

\paragraph{Visual stimulus.}
The visual stimulus consisted of full-field drifting gratings with 90\% contrast, luminance of 10 cd/m$^2$, spatial frequency of 0.08 cycles/degree, and temporal frequency of 2 cycles/s. Two types of stimuli were presented for each imaging site: First, directional tuning was mapped using a pseudo-random sequence of drifting gratings at sixteen directions of motion, 500 ms per direction, without blanks, with 12--30 trials for each direction of motion.  Second, ro measure correlations, the stimulus was modified to include only two directions of motion (in 9 datasets) or five directions (in 22 datasets) and the gratings were presented for 1 second and were separated by 1-second blanks, with 100--300 trials for each direction of motion.

@@

The measured fluorescent traces were deconvolved to reconstruct the firing rates for each neuron: First, the first principal component was subtracted from the raw traces in order to reduce common mode noise related to small cardiovascular movements \cite{Cotton:2013}. The resulting traces were high-pass filtered above 0.1 Hz and downsampled to 20 Hz (\figref{2}{C}). Then, the firing rates were estimated using by nonnegative deconvolution \cite{Vogelstein:2010}.

Orientation tuning was computed by fitting the mean firing rates for each direction of motion $\phi$ using two-peaked von Mises tuning functions $f(\phi)=a + b\exp\left[\frac 1 w(\cos(\phi-\theta)-1) \right] + c\exp\left[\frac 1 w(\cos(\phi-\theta+\pi)-1) \right]$ where $b\ge c$ are the amplitudes of the two respective peaks, $w$ is the tuning width, and  $\theta$ is the preferred direction. The significance of the fit was determined by the permutation test: the labels of the direction were randomly permuted 10,000 times; the $p$-values of the fits were computed as the fraction of permutations that yielded $R^2$ equal to or higher than that of the unpermuted data.  Cells were considered tuned with $p<0.05$.

@@

\paragraph{Cross-validation.}
To compare the performance of the estimators, we used conventional 10-fold cross-validation: Trials were randomly divided into 10 subsets with approximately equal numbers of trials of each condition in each subset. Each subset was then used as the testing sample with the rest of the data used as the training sample for estimating the covariance matrix. The average validation loss over the 10 folds was reported.

@@

When the validation loss was not required, only the inner loop of cross-validation was used on the entire dataset.  This approach was used to compute the covariance matrix estimates and their excess loss in the simulation study (\figref{1}{\,Rows 4 and 5}) and to analyze the partial correlation structure of the $C_{\sf sparse+latent}$ estimator (Fig.~\ref{fig:4}--\ref{fig:6}).

@@

Estimator $C_{\sf diag}$ (Eq.~\ref{eq:c-diag})  used two hyperparameters: the covariance shrinkage intensity $\lambda \in [0,1]$ and variance shrinkage intensity $\alpha \in [0,1]$.  The variances (the diagonal of $C_{\sf sample}$) were shrunk linearly toward their mean value $\frac 1 p \Tr(C_{\sf sample})$:
\begin{equation}
D = (1-\alpha)C_{\sf sample}\circ I + \alpha \frac 1 p \Tr(C_{\sf sample}) I
\end{equation}
The $C_{\sf diag}$ estimate was then obtained by shrinking $C_{\sf sample}$ toward $D$ according to Eq.~\ref{eq:c-diag}.

In estimator $C_{\sf factor}$ (Eq.~\ref{eq:c-factor}), the low-rank matrix $L$ and the diagonal matrix $D$ are found by solving the minimization problem
\begin{equation}
(L,D) = \argmin\limits_{\hat L,\hat D} \loss{\hat L + \hat D,C_{\sf sample}},
\end{equation}
by an expectation-maximization (EM) algorithm with specified rank of $L$. Additional the diagonal matrix of individual variances is shrunk toward its mean value according to Eq.~\ref{eq:c-factor}.

In estimator $C_{\sf sparse}$ (Eq.~\ref{eq:c-sparse}), the sparse precision matrix $S$ is found by minimizing the $L_1$-penalized loss with regularization parameter $\lambda$:
\begin{equation}
S = \argmin\limits_{\hat S \succ 0} \loss{{\hat S}^{-1},C_{\sf sample}} + \lambda \|\hat S \|_1
\end{equation}
where $\hat S\succ 0$ denotes the constraint that $\hat S$ be a positive-definite matrix and $\|\hat S\|_1$ is the element-wise $L_1$ norm of the matrix $\hat S$. This problem formulation is known as \emph{graphical lasso} \cite{Meinshausen:2006, Friedman:2008}. To solve this minimization problem, we adapted the alternative-direction method of multipliers (ADMM) \cite{Ma:2013}. 
Unlike $C_{\sf diag}$ and $C_{\sf factor}$, this estimator does not include linear shrinkage: the selection of the sparsity level provides sufficient flexibility to fine-tune the regularization level.

Estimator $C_{\sf sparse+latent}$ (Eq.~\ref{eq:c-sl}) estimates a larger sparse precision matrix $S^\ast$ of the joint distribution of the $p$ observed neurons and $d$ latent units.  
\begin{equation}
S^\ast=
\begin{pmatrix}
S & S_{12} \\
S_{12}^\T & S_{22}
\end{pmatrix},
\end{equation}
where the $p\times p$ partition $S$ corresponds to the visible units. 
Then the covariance matrix of the observed population is 
\begin{equation}
C_{\sf sparse+latent} = \left(S-S_{12}S_{22}^{-1}S_{12}^\T\right)^{-1}
\end{equation}
The rank of the $p\times p$  matrix $L=S_{12}S_{22}^{-1}S_{12}^\T$ matches the number of the latent units in the joint distribution. Rather than finding $S_{12}$ and $S_{22}$ separately, $L$ can be estimated as a low-rank matrix. We adapted the ADMM algorithm to minimize the loss function with $L_1$ penalty on $S$ to regulate its sparseness and nuclear norm penalty on $L$ to regulate its rank \cite{Chandrasekaran:2010,Ma:2013}:
\begin{equation}\label{eq:ma}
(S,L) = \argmin\limits_{\hat S,\hat L} \loss{\hat S-\hat L, C_{\sf sample}} + \alpha\|\hat S\|_1 + \beta\Tr(\hat L)
\end{equation}

The partial correlation matrix (Eq.~\ref{eq:partial}) computed from $C_{\sf sparse+latent}$ includes interactions between the visible and latent units and was used \figref{4}{C and D} and \figref{5}{C}, and \figref{6}{A--C}).  The partial correlation matrix computed from $S$ alone expresses strengths of pairwise interactions 
\begin{equation}
P_{\sf sparse} = -(S\circ I)^{-\frac 1 2} S  (S\circ I)^{-\frac 1 2}
\end{equation}
and were used in \figref{4}{F, G, H}.

@@

\paragraph{Cross-validation with conditioned variances.}
This study required an unbiased estimate of the discrepancy between various correlation matrix estimates and the true value of the correlation matrix. This was accomplished by cross-validation whereby regularized covariance matrix estimates computed from training samples were compared to unbiased covariance matrix estimates computed from independent testing samples. Loss functions were defined with respect to the covariance matrix rather than the correlation matrix because no unbiased estimate of the correlation matrix exists for finite sample sizes in the general case \cite{Fisher:1921}. In particular, we showed that with the normal loss function (Eq.~\ref{eq:loss}), the validation loss $\loss{C,C_{\sf sample}^\prime}$ is an unbiased estimate of true loss when $C_{\sf sample}^\prime$ is an unbiased estimate. However, this approach only worked under the assumption of uniform variances across all conditions included in the training and testing samples.  Here, we extend the validation loss to the estimation and evaluation of a common correlation matrix across multiple conditions with different variances.

Let $T_c$ and $T_c^\prime$ denote the time bin indices for the training and testing samples, respectively, limited to condition $c$. Here, the prime symbol marks quantities estimated from the testing sample.

@@

Note that $\bar x_c= \frac 1 {n_c} \sum\limits_{t \in T_c}x(t)$ is estimated from the training sample but used in both estimates, making $C_{c,\sf sample}^\prime$ an unbiased estimate of the true covariance matrix, $\Sigma$. As such, $C_{c,\sf sample}^\prime$ can be used for validation.

The common correlation matrix $R_{\sf sample}$ is estimated by averaging the condition-specific correlations:
\begin{equation}
	R_{\sf sample}  
	= \frac 1 n \sum\limits_c n_c \left(V_{c,\sf sample}^{-\frac 1 2}C_{c,\sf sample}V_{c,\sf sample}^{-\frac 1 2}\right)
        = \frac 1 n \sum\limits_c \sum\limits_{t \in T_c} z(t)z(t)^\T,
\end{equation}
where $n=\sum\limits_c n_c$ and $V_{c,\sf sample} = C_{c,\sf sample} \circ I$ is the diagonal matrix containing the sample variances. Then $R_{\sf sample}$ is simply the covariance matrix of the $z$-score signal 
$z(t) = V_{c,\sf sample}^{-\frac 1 2} \left(x(t) - \bar x_c\right)$ of the training sample.
